Using Optimal Time Step Selection to Boost the Accuracy of Finite Difference Schemes for Time Dependent PDEs

### 2010-05-02 papers/OTSPDE/OTSPDE.tex
Revised OTSPDE paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First completed draft: 2008/05/12, Vitamin D, Inc.
% Submission draft: 2008/06/08, Vitamin D, Inc.
% Resubmission draft: 2008/11/16, Vitamin D, Inc.
% Resubmission draft: 2009/05/25, Serendipity Research
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{Boosting the Accuracy of Finite Difference Schemes via
       Optimal Time Step Selection and Non-Iterative Defect Correction}

\author{Kevin T. Chu\address{Institute of High Performance Computing, A*STAR, Singapore, Singapore}$^,$\address{Vitamin D, Inc., Menlo Park, CA 94025, United States}
}

\runtitle{Optimal Time Step Selection and Non-Iterative Defect Correction for 
          FD Schemes}
\runauthor{K.T. Chu}

@@

\begin{abstract}
In this article, we present a simple technique for boosting the order of 
accuracy of finite difference schemes for time dependent 
partial differential equations by optimally selecting the time step used
to advance the numerical solution and adding defect correction terms in a
non-iterative manner.  The power of the technique is its ability to extract 
as much accuracy as possible from existing finite difference schemes with 
minimal additional effort.  Through straightforward numerical analysis 
arguments, we explain the origin of the boost in accuracy and estimate the 
computational cost of the resulting numerical method.  We demonstrate the 
utility of optimal time step (OTS) selection combined with non-iterative 
defect correction (NIDC) on several different types of finite difference
schemes for a wide array of classical linear and semilinear PDEs in one and 
more space dimensions on both regular and irregular domains.  
\end{abstract}

@@

\section{Introduction}
High-order numerical methods for partial differential equations (PDEs) will 
always be valuable for increasing the computational efficiency of numerical 
simulations.  Thus, it is not at all surprising that a great deal of effort in 
numerical PDEs continues to be focused on the development of high-order 
numerical 
schemes~\cite{spotz_2001,bruger_2005,gibou_2005,ito_2005,shukla_2005,heidenreich_2007,shukla_2007}.  
Typically, high-order accuracy is achieved by constructing
schemes that are formally high-order accurate.  However, high-order 
accuracy can also be obtained by using formally low-order schemes in clever 
ways (\eg Richardson extrapolation and compact finite difference 
schemes~\cite{spotz_2001}).  When possible, the latter approach can be a 
powerful way to boost the accuracy of a numerical method \emph{without} 
introducing too much additional algorithmic (and programming) complexity.

Optimization of numerical parameters (\eg $\dx$ and $\dt$) is an important, 
but relatively uncommon, approach for maximizing the accuracy of numerical 
schemes.  Often, all combinations of numerical parameters are considered 
equally good.  For some problems, however, optimizing the numerical parameters 
can significantly increase the accuracy of the numerical solution.  For 
example, the well-known unit CFL condition, $\dt = \dx/a$, for the first-order 
upwind forward Euler discretization of the linear advection equation 
completely eliminates the numerical error introduced by the 
scheme~\cite{leveque_book_2002}.  Another well-known example is the boost in 
the order of accuracy that results from setting the time step to 
$\dt = \dx^2/6D$ when the 1D diffusion equation is solved using forward Euler 
time integration with the standard second-order central difference 
approximation for the Laplacian.  As a final example, Zhao showed that an 
appropriate choice of the weight factor $\theta$ in the theta-method can 
significantly decrease the error in finite element solutions to the diffusion 
equation~\cite{zhao_2006}.

In this article, we present a simple technique for boosting the order of 
accuracy of finite difference schemes for time dependent PDEs by optimally 
choosing the time step and, when necessary, adding defect correction terms
in a \emph{non-iterative} manner.  This technique is a systematic 
generalization of the ideas underlying some of the examples mentioned above.
Optimal time step (OTS) selection is based on the observation that a carefully 
chosen time step can simultaneously eliminate low-order terms in both the
spatial and temporal the discretization errors.  It combines the well-known 
approach of using knowledge of the structure of the PDE to reduce numerical 
error with the less-common notion that an optimal choice of numerical 
parameters can boost the effectiveness of algorithmically simple and 
computationally low-cost numerical schemes.  For some problems, OTS alone is 
sufficient to allow a formally low-order method to deliver high-order 
accuracy.  However, addition of defect correction terms is generally required 
to eliminate residual terms in the leading-order truncation error.  It is 
important to emphasize that, unlike traditional defect/deferred correction 
methods~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006} 
which use defect correction terms to iteratively refine the solution at 
each time step, we take a non-iterative defect correction (NIDC) approach
that incorporates defect correction terms directly into the original finite 
difference scheme.

As a technique for enhancing numerical methods, OTS-NIDC has several desirable 
features.  Like any technique that leads to high-order methods, OTS-NIDC
produces schemes that greatly reduce the computational cost required to obtain 
a numerical solution of a desired accuracy.  However, its real power comes 
from the fact that it achieves high-order accuracy while being based solely on 
very simple, formally low-order finite difference schemes.  In other words, 
OTS-NIDC allows us to extract as much accuracy as possible from an existing 
numerical scheme with minimal additional work.  For instance, we will see how 
OTS-NIDC makes it possible to achieve 4th-order accuracy for the diffusion 
equation in any number of space dimensions using only a second-order stencil 
for the Laplacian and explicit time integration.  
Moreover, high-order accuracy is not restricted to problems 
on simple, rectangular domains; irregular domains are straightforward to 
handle by appropriately setting the values in ghost 
cells~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
OTS-NIDC is even beneficial when using finite difference schemes to solve some 
nonlinear PDEs.  While there are certainly limitations to OTS-NIDC, we will 
see that there are many problems where it is useful.

This article is organized as follows.  In the remainder of this section,
we compare OTS-NIDC to a few related numerical techniques.
In Section~\ref{sec:OTS-NIDC}, we present the theoretical foundations of 
OTS-NIDC and show how it boosts the order of accuracy for formally low-order 
numerical schemes.  
In Section~\ref{sec:applications_1d}, we demonstrate the utility of 
OTS-NIDC to problems in one space dimension by applying it to finite 
difference schemes for several classical linear and nonlinear PDEs.  
In Section~\ref{sec:applications_multidim}, we show that 
OTS-NIDC is useful for problems in multiple space dimensions by applying it
the 2D linear advection equation and the 2D diffusion equation.  To illustrate 
the ease with which high-order accuracy can be achieved for problems on 
arbitrary domains, the 2D diffusion equation is solved on both regular 
and irregular domains.  
For each PDE in Sections~\ref{sec:applications_1d} and 
\ref{sec:applications_multidim}, we provide the numerical analysis required 
to calculate the optimal time step size, the defect correction terms, and the 
order of accuracy.  Finally, we offer some concluding remarks in 
Section~\ref{sec:summary}.

\subsection{Comparison with Traditional Defect/Deferred Correction Methods}
Defect/deferred correction is a well-known strategy for boosting the order of 
accuracy of finite difference schemes for time dependent PDEs.  It usually 
takes one of two forms: 
(1) iterative refinement of the numerical solution~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}
or 
(2) derivation of high-order compact schemes for the spatial derivative 
operator treating time derivatives as source 
terms~\cite{spotz_2001,ito_2005,heidenreich_2007}.

For iterated defect/deferred correction, the numerical solution at each 
time step is computed by iteratively adding correction terms that are 
calculated from lower accuracy iterates.  When used to improve the 
temporal order of accuracy, the iterative approach can help to ensure 
stability of the time integration scheme~\cite{kress_2006}.  Although there 
are instances where iterative defect correction has been used to eliminate 
both spatial and temporal errors~\cite{gustafsson_2002}, it is usually used to 
improve only one of these errors~\cite{pereyra_1968,kress_2002,kress_2006}.
Note that unlike iterated defect correction, OTS-NIDC requires no iteration 
during each time step because it computes correction terms using only the 
solution at the previous time step(s) (as opposed to requiring an estimate 
of the solution at the current time).

@@

OTS-NIDC has several advantages over both of these defect correction 
techniques.  First, OTS-NIDC is generally more computationally efficient than 
both iterated defect correction and the high-order compact scheme approach
because it requires less memory and, for many problems, admits simple 
explicit time integration schemes.  Second, OTS-NIDC inherently controls both 
spatial and temporal numerical errors at the same time.  Finally, OTS-NIDC 
does not rely solely on the addition of finite differences of solution values 
to cancel out discretization errors.  Instead, it takes advantage of 
relationships between spatial derivatives, temporal derivatives and source 
terms to eliminate the need for finite differences approximations of 
high-order derivatives of the solution.  Compared with iterated defect 
correction, OTS-NIDC makes it easier to avoid instabilities that arise from 
the presence of finite difference discretizations of high-order temporal 
derivatives and may eliminate some of the correction terms.  Comparing to the
high-order compact scheme approach, the optimization of numerical parameters 
in OTS-NIDC means that mixed partial derivatives do not need to appear in the 
fully discrete equations.

@@

\section{\label{sec:OTS-NIDC}
         Theoretical Foundations of OTS Selection and NIDC}
OTS-NIDC is not by itself a method for constructing finite difference 
schemes.  Rather, it is a way to enhance to the performance of existing 
finite difference schemes by carefully choosing the time step and adding
defect correction terms to eliminate low-order numerical errors.  
The two key ideas underlying OTS-NIDC are 
(1) a judicious choice of time step can be used to eliminate the leading-order 
(and possibly higher-order) terms in the error
and
(2) the PDE provides valuable insight into the relationships between terms
in the discretization error. 
In this section, we present and analyze the technique of OTS-NIDC for finite 
difference schemes constructed to solve scalar evolution equations of the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important, 
restricting our attention to problems of this form is not a serious 
limitation.  The same principles apply to finite difference schemes for 
general PDEs.

\subsection{\label{sec:ots_model_pde} 
            Optimal Time Step for a Model PDE}
Let us begin by considering a finite difference approximation for the
following time dependent PDE in one spatial dimension: 
\beq
  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n},
  \label{eq:model_PDE}
\eeq
where $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
is well-posed.  Because we will be boosting the accuracy of the finite 
difference approximation by optimizing the time step, there is no reason to 
explicitly construct the scheme to be high-order.  In fact, using a high-order 
finite difference scheme only complicates the analysis required to calculate 
the optimal time step.

@@

An important class of finite difference schemes for time dependent PDEs are 
those based on the method of lines.   When the method of lines approach is 
used for (\ref{eq:model_PDE}), $P$ often takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
discretization errors, and $\alpha$ and $\beta$ are constants that 
depend on the details of the PDE and finite difference scheme.  Setting 
$P = 0$, we find that the optimal time step is given by\footnote{The $\dt = 0$ 
root of $P$ is discarded because it is not numerically meaningful.}
\beq
  \dto = \left(\alpha/\beta \right)^{1/s} \dx^{r/s}.
  \label{eq:optimal_time_step}
\eeq
Notice that the optimal time step has a similar functional dependence on the 
grid spacing as a time step that satisfies a stability constraint.

@@

\subsubsection*{\label{sec:computational_performance} 
                Computational Performance}
For many problems, the optimal time step is on the order of the maximum 
time step allowed by the stability constraint for an explicit method.  
As a result, using the optimal time step to advance the finite
difference scheme in time may appear to be overly restrictive.
However, this apparent drawback is more than 
compensated by the boost in the order of accuracy.  For example, when solving 
the diffusion equation using a forward Euler time integration scheme with a 
second-order central difference stencil for the Laplacian, we obtain a 
fourth-order scheme when $\dt$ is set equal to the optimal time step 
(discussed further in Sections~\ref{sec:diffusion_eqn_1d} and
\ref{sec:diffusion_eqn_2d}).  
As Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
scheme that is computationally cheaper than traditional methods that use 
the same low-order finite difference stencils.  The performance gain for 
problems in higher spatial dimensions is even more impressive.  As we can see 
in Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
simple forward Euler scheme with the optimal time step and the Crank-Nicholson
method grows with the spatial dimension of the problem.

\begin{table}[tb]
\caption{
Computational cost of various finite difference schemes for the 1D 
diffusion equation as a function of the global numerical error $e$.
For all of the schemes, the standard second-order central difference 
stencil is used to discretize the Laplacian.  
Note that the time step for the backward Euler scheme is chosen so that the
temporal error is subdominant to the spatial error and that the computation 
time for the Crank-Nicholson method assumes an $O(N) = O \left( 1/\dx\right)$ 
matrix inversion algorithm. 
}
\label{tab:comp_perf_vs_err} 
\renewcommand{\arraystretch}{1.5}
\centering
\begin{tabular}{lcccc}
  \hline
  {\bf Numerical Scheme} & $\dx$ 
  & $\dt$
  & {\bf Memory}
  & {\bf Compute Time}
  \\
  \hline 
  Forward Euler    & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Backward Euler   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Crank-Nicholson  & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-1} \right)$ \\
  Forward Euler with OTS-NIDC  & $O\left( e^{1/4} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/4} \right)$ 
                   & $O\left( e^{-3/4} \right)$ \\ 
  \hline
\end{tabular}
\end{table}

\begin{table}[tb]
\caption{
Computational cost of the forward Euler scheme with OTS-NIDC and
the Crank-Nicholson method for solving the diffusion equation as a function 
of the global numerical error $e$.
Note that the computation time for the Crank-Nicholson method assumes an 
$O(N) = O \left( 1/\dx^d \right)$ matrix inversion algorithm. 
A comparison of computational costs in higher spatial dimensions continues 
along the same trends but requires replacement of the forward Euler time 
integration with an alternative for which the optimal time step is stable 
(\eg the DuFort-Frankel scheme). 
}
\label{tab:comp_perf_vs_dim}
\renewcommand{\arraystretch}{1.5}
\centering
\begin{tabular}{ccccc}
  \hline
  & \multicolumn{2}{c}{\bf Forward Euler with OTS-NIDC} 
  & \multicolumn{2}{c}{\bf Crank-Nicholson} \\
  \cline{2-3} \cline{4-5} 
    {\bf Dimensions} & {\bf Memory} & {\bf Compute Time} 
  & {\bf Memory} & {\bf Compute Time} \\
  \hline 
  $1$ & $O\left( e^{-1/4} \right)$ 
      & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ \\ 
  $2$ & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-3/2} \right)$ \\ 
  $3$ & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-5/4} \right)$ 
      & $O\left( e^{-3/2} \right)$ 
      & $O\left( e^{-2} \right)$ \\
  \hline 
\end{tabular}
\end{table}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First completed draft: 2008/05/12, Vitamin D, Inc.
% Submission draft: 2008/06/08, Vitamin D, Inc.
% Resubmission draft: 2008/11/16, Vitamin D, Inc.
% Resubmission draft: 2009/05/25, Serendipity Research
% Resubmission draft: 2010/03/25, Serendipity Research
% Resubmission draft: 2010/05/02, Serendipity Research
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{Using Optimal Time Step Selection to Boost the Accuracy of
       Finite Difference Schemes for Time Dependent PDEs}

\author{Kevin T. Chu\address{Serendipity Research, Mountain View, CA 94041, United States}
}

\runtitle{OTS Selection for FD Approximations of Time Dependent PDEs}
\runauthor{K.T. Chu}

@@

\begin{abstract}
In this article, we present a simple technique for boosting the order of 
accuracy of finite difference schemes for time dependent 
partial differential equations by optimally selecting the time step used
to advance the numerical solution. The power of the technique is its ability
to increase the accuracy of existing finite difference schemes with 
minimal modification to the original numerical algorithm.  Through
straightforward numerical analysis arguments, we explain the origin of the
boost in accuracy and estimate the computational cost of the resulting
numerical method.  We demonstrate the utility of optimal time step (OTS)
selection on several different types of finite difference schemes
for a wide array of classical linear and semilinear PDEs in one and more space
dimensions on both regular and irregular domains.  
\end{abstract}

@@

\section{Introduction}
High-order numerical methods for partial differential equations (PDEs) will 
always be valuable for increasing the computational efficiency of numerical 
simulations.  Thus, it is not at all surprising that a great deal of effort in 
numerical PDEs continues to be focused on the development of high-order 
numerical 
schemes~\cite{spotz_2001,bruger_2005,gibou_2005,ito_2005,shukla_2005,heidenreich_2007,shukla_2007}.  
Typically, high-order accuracy is achieved by constructing
schemes that are formally high-order accurate.  However, high-order 
accuracy can also be obtained by using formally low-order schemes in clever 
ways (\eg Richardson extrapolation and compact finite difference 
schemes~\cite{spotz_2001}).  When possible, the latter approach can be a 
powerful way to boost the accuracy of a numerical method without
introducing too much additional algorithmic (and programming) complexity.

One important, but relatively uncommon, approach for increasing the accuracy
of a numerical algorithm is optimization of its parameters
(\eg $\dx$ and $\dt$).  Often, all combinations of numerical parameters are
considered equally good.  For some problems, however, optimizing the numerical
parameters can dramatically increase the accuracy of the numerical solution.
For example, the well-known unit CFL condition, $\dt = \dx/a$, for the
first-order upwind forward Euler discretization of the linear advection
equation completely eliminates the numerical error introduced by the 
scheme~\cite{leveque_book_2002}.  Another relatively well-known example is the
boost in the order of accuracy that results from setting the time step to 
$\dt = \dx^2/6D$ when the 1D diffusion equation is solved using forward Euler 
time integration with the standard second-order central difference 
approximation for the Laplacian.  As a final example, Zhao recently showed
that an appropriate choice of the weight factor $\theta$ in the theta-method
can significantly decrease the error in finite element solutions to the
diffusion equation~\cite{zhao_2006}.

In this article, we present a simple technique for boosting the \emph{order} of 
accuracy of finite difference schemes for time dependent PDEs by optimally 
choosing the time step.  This technique is a systematic generalization of
the ideas underlying some of the examples mentioned above.  Optimal time step
(OTS) selection is based on the observation that a carefully chosen time step
can simultaneously eliminate low-order terms in both the spatial and temporal
the discretization errors.  At its essence, OTS selection combines the 
powerful but uncommon approach of optimally choosing numerical parameters 
with the well-known approach of using knowledge of the structure of the
PDE to reduce numerical error.  The end result is a strategy for boosting
the order of accuracy of algorithmically simple and computationally low-cost
numerical schemes.

While OTS selection alone is sufficient to allow a formally low-order method to
deliver high-order accuracy for some problems, defect correction terms are
generally required to eliminate residual terms in the leading-order truncation
error.  However, unlike traditional defect correction 
methods~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}, 
using OTS selection eliminates the need to incorporate correction terms that
involve derivatives of higher-order than those that appear in the original PDE.
Another significant feature of OTS selection with defect correction (OTS-DC)
is that the defect terms are incorporated in a \emph{non-iterative} fashion.
In contrast, traditional defect/deferred correction approaches typically
require several refinement iterations at each time step to boost the order of
accuracy of the numerical solution.  OTS-DC also differs from the method of
modified equations~\cite{griffiths_1986} for eliminating higher-order
corrections to the truncation error because it uses an optimal time
step, not explicit inclusion of correction terms, to regularize the
singularly perturbed modified equations - the correction terms that remain
when using OTS-DC are all regular perturbations to the original equation.

As a technique for enhancing numerical methods, OTS selection has several
desirable features.  Like any technique that leads to high-order methods,
OTS selection produces schemes that greatly reduce the computational cost
required to obtain a numerical solution of a desired accuracy.  However, its
real power comes from the fact that it achieves high-order accuracy while
being based solely on very simple, formally low-order finite difference
schemes \emph{without} requiring special algorithmic steps.  In other words,
OTS selection allows us to extract as much accuracy as possible from an
existing numerical scheme with minimal additional work.  For instance, we will
see how OTS selection makes it possible to achieve 4th-order accuracy for the
diffusion equation in any number of space dimensions using only a second-order
stencil for the Laplacian and explicit time integration.  Moreover, high-order
accuracy is not restricted to problems on simple, rectangular domains;
irregular domains are straightforward to handle by appropriately setting the
values in ghost cells~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
OTS selection is even beneficial when using finite difference schemes to solve
some nonlinear PDEs.

This article is organized as follows.  In the remainder of this section,
we compare OTS selection to a few related numerical techniques.
In Section~\ref{sec:OTS}, we present the theoretical foundations of 
OTS selection and show how it boosts the order of accuracy for formally
low-order numerical schemes.  
In Section~\ref{sec:applications_1d}, we demonstrate the utility of 
OTS selection to problems in one space dimension by applying it to finite 
difference schemes for several classical linear and nonlinear PDEs.  
In Section~\ref{sec:applications_multidim}, we show that 
OTS selection is useful for problems in multiple space dimensions by applying
it the 2D linear advection equation and the 2D diffusion equation.  To
illustrate the ease with which high-order accuracy can be achieved for
problems on arbitrary domains, the 2D diffusion equation is solved on both
regular and irregular domains.  For each PDE in
Sections~\ref{sec:applications_1d} and \ref{sec:applications_multidim}, we
provide the numerical analysis required to calculate the optimal time step
size, the defect correction terms, and the order of accuracy.  Note that in
order to emphasize the method, we have chosen to apply it to relatively simple
problems rather than use it to tackle ``real-life'' applications.  The method,
however, is not restricted to these simple examples.  Finally, in
Section~\ref{sec:summary}, we summarize the paper with some concluding
remarks.

\subsection{Comparison with Defect/Deferred Correction Methods}
Defect/deferred correction is a well-known strategy for boosting the order of 
accuracy of finite difference schemes for time dependent PDEs.  It usually 
takes one of two forms: 
(1) iterative refinement of the numerical
solution~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}
or 
(2) derivation of high-order compact schemes for the spatial derivative 
operator treating time derivatives as source 
terms~\cite{spotz_2001,ito_2005,heidenreich_2007}.

For iterated defect/deferred correction, the numerical solution at each 
time step is computed by iteratively adding correction terms that are 
calculated from lower accuracy iterates.  When used to improve the 
temporal order of accuracy, the iterative approach helps to ensure 
stability of the time integration scheme~\cite{kress_2006}.  Although there 
are instances where iterative defect correction has been used to eliminate 
both spatial and temporal errors~\cite{gustafsson_2002}, it is usually used to 
improve only one of these errors~\cite{pereyra_1968,kress_2002,kress_2006}.
Unlike iterated defect correction, OTS-DC requires no iteration during each
time step because it computes correction terms using only the solution at the
previous time step(s) (as opposed to requiring an estimate of the solution at
the current time).

@@

OTS-DC has several advantages over both of these defect correction 
techniques.  First, OTS-DC is generally more computationally efficient than 
both iterated defect correction and the high-order compact scheme approach
because it requires less memory and, for many problems, admits simple 
explicit time integration schemes.  Second, OTS-DC inherently controls both 
spatial and temporal numerical errors at the same time.  Finally, taking
advantage of relationships between spatial derivatives, temporal derivatives
and source terms, OTS-DC avoids the need to explicitly incorporate finite
differences of high-order derivatives of intermediate solutions in order to
eliminate discretization errors. 
Compared with iterated defect correction, OTS-DC makes it easier to avoid
instabilities that arise from the presence of finite difference discretizations
of high-order temporal derivatives by eliminating the most problematic
correction terms.  Comparing to the high-order compact scheme approach, the
optimization of numerical parameters in OTS-DC means that mixed partial
derivatives do not need to appear in the fully discrete equations.

\subsection{Comparison with the Modified Equation Method}
Both defect and deferred correction can be viewed as special cases of the
method of modified equations~\cite{griffiths_1986}.  When using any modified
equation method, the basic approach is to derive a modified version of the 
original PDE through the introduction of the leading-order terms of the
discretization error into the PDE.  Unfortunately, the modified PDE is a
singular perturbation to the original PDE due to the presence of the
discretization parameters in the correction terms.  Because the the singular
perturbation can lead to numerical instabilities, finding a way regularize
the modified equation is an important step for developing any numerical method
based on the modified equations method.

Whereas deferred correction methods deal with the singular perturbation by
using an iterative procedure to avoid numerical instabilities, the traditional
method of modified equations~\cite{griffiths_1986} handles the
singularity by ``pushing'' the singularity to a higher-order term in the
discretization error.  The basic procedure involves substituting derivatives
that appear in the singular term with expressions obtained by taking
derivatives of the original PDE.  This procedure yields a new modified
equation where the new singular term is of sufficiently high-order in the
discretization parameter that it can be neglected in the construction of the
numerical scheme.

Like defect/deferred correction, OTS selection (even without defect correction)
falls within the general framework of the method of modified equations.
In OTS selection, regularization of the singular perturbation in the modified
equations is achieved by choosing the time step so that the spatial and
temporal singular perturbation terms exactly cancel each other.  More
specifically, by constraining the relationship between the time step and
grid spacing, the modified equation becomes a regular perturbation to the
original PDE.  In the special case when OTS selection does not require
defect correction, the modified equation is actually exactly the same as the
original equation -- no regular perturbation terms remain after the singular
perturbation terms have been eliminated.

There are several advantage of OTS selection over traditional modified
equation methods.  First, because no finite difference approximations of
higher-order derivatives are required, numerical schemes based on OTS
selection do not require additional boundary conditions.  Second, derivation
of the optimal time step and correction terms for OTS-DC requires a comparable
amount of analysis as the traditional modified equation method.  In all cases,
we need explicit knowledge of the leading-order terms in the discretization
error.  Finally, because the leading-order terms in the discretization error
are eliminated implicitly rather than explicitly, the computational
performance of OTS-DC is typically superior to the performance of traditional
modified equation methods.

@@

\section{\label{sec:OTS} Theoretical Foundations of OTS Selection}
OTS selection is not by itself a method for constructing finite difference 
schemes.  Rather, it is a way to enhance to the performance of existing 
finite difference schemes by carefully choosing the time step to eliminate
leading-order terms in the discretization error.  The two key ideas
underlying OTS selection are 
(1) a judicious choice of time step can be used to eliminate the leading-order 
(and possibly higher-order) terms in the error
and
(2) the PDE provides valuable insight into the relationships between terms
in the discretization error. 
In this section, we present and analyze the technique of OTS selection for
finite difference schemes constructed to solve scalar evolution equations of
the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important, 
restricting our attention to problems of this form is not a serious 
limitation.  The same principles apply to finite difference schemes for 
general PDEs.

\subsection{\label{sec:ots_model_pde} 
            Optimal Time Step Selection for a Model PDE}
Let us begin by considering a finite difference approximation for the
following time dependent PDE in one spatial dimension: 
\beq
  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n},
  \label{eq:model_PDE}
\eeq
where $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
is well-posed.  Because we will be boosting the accuracy of the finite 
difference approximation by optimizing the time step, there is no reason to 
explicitly construct the scheme to be high-order.  In fact, using a high-order 
finite difference scheme only complicates the analysis required to calculate 
the optimal time step.

@@

An important class of finite difference schemes for time dependent PDEs are 
those based on the method of lines.   When the method of lines approach is 
used for (\ref{eq:model_PDE}), $P$ often takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
discretization errors, and $\alpha$ and $\beta$ are constants that 
depend on the details of the PDE and finite difference scheme.  Setting 
$P = 0$, we find that the optimal time step is given by\footnote{The $\dt = 0$ 
root of $P$ is discarded because it is not numerically meaningful.}
\beq
  \dto = \left(\alpha/\beta \right)^{1/s} \dx^{r/s}.
  \label{eq:optimal_time_step}
\eeq
Notice that the optimal time step has a similar functional dependence on the 
grid spacing as a time step constraint required for stability of an explicit
time-stepping scheme.

@@

\subsubsection*{\label{sec:computational_performance} 
                Computational Performance}
For many problems, the optimal time step is on the order of the maximum 
time step allowed by the stability constraint for an explicit method.  
As a result, using the optimal time step to advance the finite
difference scheme in time may appear to be overly restrictive.
However, this apparent drawback is more than 
compensated by the boost in the order of accuracy.  For example, when solving 
the diffusion equation using a forward Euler time integration scheme with a 
second-order central difference stencil for the Laplacian, we obtain a 
fourth-order scheme when $\dt$ is set equal to the optimal time step 
(discussed further in Sections~\ref{sec:diffusion_eqn_1d} and
\ref{sec:diffusion_eqn_2d}).  
As Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
scheme that is computationally cheaper than traditional implicit schemes that
use the same low-order finite difference stencils.  The performance gain for 
problems in higher spatial dimensions is even more impressive.  As we can see 
in Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
simple forward Euler scheme with the optimal time step and the Crank-Nicholson
method grows with the spatial dimension of the problem.

\begin{table}[tb]
\caption{
Computational cost of various finite difference schemes for the 1D 
diffusion equation as a function of the global numerical error $e$.
For all of the schemes, the standard second-order central difference 
stencil is used to discretize the Laplacian.  
Note that the time step for the backward Euler scheme is chosen so that the
temporal error is subdominant to the spatial error and that the computation 
time for the Crank-Nicholson method assumes an $O(N) = O \left( 1/\dx\right)$ 
matrix inversion algorithm. 
}
\label{tab:comp_perf_vs_err} 
\renewcommand{\arraystretch}{1.5}
\centering
\begin{tabular}{lcccc}
  \hline
  {\bf Numerical Scheme} & $\dx$ 
  & $\dt$
  & {\bf Memory}
  & {\bf Compute Time}
  \\
  \hline 
  Forward Euler    & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Backward Euler   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-3/2} \right)$ \\
  Crank-Nicholson  & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/2} \right)$ 
                   & $O\left( e^{-1} \right)$ \\
  Forward Euler with OTS-DC & $O\left( e^{1/4} \right)$ 
                   & $O\left( e^{1/2} \right)$ 
                   & $O\left( e^{-1/4} \right)$ 
                   & $O\left( e^{-3/4} \right)$ \\ 
  \hline
\end{tabular}
\end{table}

\begin{table}[tb]
\caption{
Computational cost of the forward Euler scheme with OTS-DC and
the Crank-Nicholson method for solving the diffusion equation as a function 
of the global numerical error $e$.
Note that the computation time for the Crank-Nicholson method assumes an 
$O(N) = O \left( 1/\dx^d \right)$ matrix inversion algorithm. 
A comparison of computational costs in higher spatial dimensions continues 
along the same trends but requires replacement of the forward Euler time 
integration with an alternative for which the optimal time step is stable 
(\eg the DuFort-Frankel scheme). 
}
\label{tab:comp_perf_vs_dim}
\renewcommand{\arraystretch}{1.5}
\centering
\begin{tabular}{ccccc}
  \hline
  & \multicolumn{2}{c}{\bf Forward Euler with OTS-DC} 
  & \multicolumn{2}{c}{\bf Crank-Nicholson} \\
  \cline{2-3} \cline{4-5} 
    {\bf Dimensions} & {\bf Memory} & {\bf Compute Time} 
  & {\bf Memory} & {\bf Compute Time} \\
  \hline 
  $1$ & $O\left( e^{-1/4} \right)$ 
      & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ \\ 
  $2$ & $O\left( e^{-1/2} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-1} \right)$ 
      & $O\left( e^{-3/2} \right)$ \\ 
  $3$ & $O\left( e^{-3/4} \right)$ 
      & $O\left( e^{-5/4} \right)$ 
      & $O\left( e^{-3/2} \right)$ 
      & $O\left( e^{-2} \right)$ \\
  \hline 
\end{tabular}
\end{table}

\subsection{\label{sec:ots_dc_general_1d_pdes} 
            OTS Selection for General PDEs in One Space Dimension} 
In the previous section, we analyzed optimal time step selection in the 
context of a simple model PDE in one space dimension.  For more general PDEs, 
OTS selection continues to be useful, but it needs to be augmented with 
defect correction (incorporated in a non-iterative manner) to achieve 
the same improvement in the order of accuracy.  In this section, we discuss 
OTS selection with defect correction (OTS-DC) in the context of general
constant coefficient, linear PDEs and semilinear PDEs.
